`r params$project_title`

### 2020-10-28 reports/rmarkdown/02_individual_reports.Rmd
adds the report for comparing microbiome datasets

---
title: "`r params$project_title`"
editor_options:
  chunk_output_type: console
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
  html_document:
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: no
params:
  organisation: Karolinska Institutet
  pi_mail: prin.cipal@ki.se
  pi_name: Prin Cipal
  project_title: Broliden_5325
  redmine_issue: '#5325'
  requester_mail: kristina.broliden@ki.se
  requester_name: Kristina Broliden
  staff_mail: supp.ortstaff@nbis.se
  staff_name: Supp Ortstaff
  staff_web: https://nbis.se/about/staff/supp-ortstaff/
---

@@

#Graph construction  
fct <- .25 # FC threshold for differential expression
pvt <- 0.05 # Pvalue threshold for differential expression
min_pct <- .1 # minimun level of detected bacteria in each sample group
```

@@

# a <- read.csv(paste0(PATH,"/ASV_tissue_B2.csv"))
# b <- read.csv(paste0(PATH,"/ASV_tissue_B1.csv"))
# rownames(a) <- paste0()
# temp <- as.matrix(a[,c((ncol(a)-6):ncol(a))])
# temp[is.na(temp)] <- "NA"
# taxa <- apply( temp ,1,function(x) paste(x,collapse ="_"))
# taxa <- sub("_NA.*","_other",taxa)

@@

#import datasets and return matrix with taxa-level resolution
datasets <- lapply(dataset_names, function(x){
  x <- read.csv( paste0(PATH,"/results/",x),row.names = 1)
  return(x)
})
names(datasets) <- dataset_names
lapply(datasets,dim)

#arrange all datasets to have the same rownames with all bacteria
# common_microbes <- unique(unlist(lapply(datasets,rownames)))
# datasets <- lapply(datasets, function(x) {
#   temp <- matrix(0,nrow = length(common_microbes),
#                  ncol = ncol(x),
#                  dimnames = list(common_microbes,colnames(x)))
#   temp[rownames(x),colnames(x)] <- x
#   return(temp)
# })
# lapply(datasets,dim)

# datasets <- datasets[2] #in case you wanna test in 1 dataset

# Transform phyloseq data into normalized matricies ( log2[CP1K] )
# datasets <- lapply(datasets,function(x){
#     x <- t(log2( t(x) / colSums(x) * 1000 + 1))
# })
# lapply(datasets,dim)
```

@@

# Create empty matrix for the full dataset
microbiome <- as.matrix(datasets[[4]])
microbiome <- microbiome[ rowSums(microbiome>0)>=2 , colSums(microbiome>0)>=2 ]

@@

```{r, eval=F,echo = FALSE}
mypar(2,2)

zscore_bac <- apply(microbiome,2,function(i){scale(i,T,T)})
zscore_bac[ zscore_bac > 6 ] <- 6
zscore_bac[ zscore_bac < -6 ] <- -6
rownames(zscore_bac) <- rownames(microbiome); colnames(zscore_bac) <- colnames(microbiome)

#Compute sample-wise correlations
cors <- cor(t(microbiome),method = "pearson")

pSNN[ pSNN < .3 ] <- 0
pSNN[ pSNN > .3 ] <- cors[pSNN > .3]
pSNN[ pSNN < 0 ] <- 0

mypar(2,2)

set.seed(1)
l <- layout_with_fr(g,niter=3000,start.temp=30)

cl_bac <- igraph::cluster_louvain(g)

plot( g , vertex.label.cex=0.000001  , 
      edge.width=  ( E(g)$weight / max(E(g)$weight)) ,layout=l,
      vertex.color= hue_pal()(length(unique(cl_bac$membership)))[factor(cl_bac$membership)],
      vertex.size=10,
      edge.color=colorRampPalette(c("grey95","black"))(90) [ round( E(g)$weight / max(E(g)$weight) * 88 )+1 ] )

legend(x = par("usr")[c(2)]+.2*diff(par("usr")[c(1,2)]),
       y = par("usr")[c(4)],
       legend = levels(factor(cl_bac$membership)),pt.bg=hue_pal()(length(unique(cl_bac$membership))),
       bty = "n",pch = 21,pt.cex = 1,xpd=T)
title("Louvain clusters",cex.main=1,font.main=1)

o_bacs <- order(cl_bac$membership)
image( t(zscore[o_bacs,o][nrow(zscore):1,]),col=heat_pal,axes=F,border=NA,
      main="bacteria",xlab="",ylab="samples",line=.4,cex.main=1,font.main=1)

text(  par("usr")[c(4)]+.01*diff(par("usr")[c(3,4)]) , seq(1,0,length.out = nrow(zscore)),
      labels = rownames(zscore), srt = 0, adj = c(0,.5), xpd = TRUE, cex=.7)
points( rep(par("usr")[c(1)]-.01*diff(par("usr")[c(1,2)]),length(cl_bac$membership)),
        (0:(length(cl_bac$membership)-1) )/(length(cl_bac$membership)-1), pch=16,
    rev((0:(length(cl$membership)-1) )/(length(cl$membership)-1)),
    col=pal[factor(cl$membership)][o_bacs] ,xpd=T)

points( (0:(length(cl$membership)-1) )/(length(cl$membership)-1),
        rep(par("usr")[c(3)]-.01*diff(par("usr")[c(3,4)]),length(cl$membership)), pch=16,
        col=pal[factor(cl$membership)][o] ,xpd=T)
points( rep(par("usr")[c(1)]-.01*diff(par("usr")[c(1,2)]),length(cl_bac$membership)), pch=16,
    rev((0:(length(cl_bac$membership)-1) )/(length(cl_bac$membership)-1)),
    col=pal[factor(cl_bac$membership)][o_bacs] ,xpd=T)

#Dataset integration

```{r, eval=F, echo = FALSE, fig.width  = 10, fig.height=6}
# Load the transcriptome datset
trx <- read.csv(paste0(PATH,"~/Downloads/TRX_rawcounts.csv"),row.names = 1)
trx <- rowsum(trx[,-c(1,2)],trx$symbol)
colnames(trx) <- sub("ML","",colnames(trx))
trx  <- trx[ rownames(trx) != "" , ]
trx <- trx[ rowSums(trx > 5) >= 3 , ]
trx <- t( log2( t( trx ) / colSums(trx) * 1e6 + 1 ) )

# Identify highly variable genes in the transcriptomic dataset
vars <- log2(apply(trx,1,var)+2)
means <- apply(trx,1,mean)
mypar(1,2)
top_variable <- names( sort(vars,decreasing = T) ) [1:3000]
plot( means , vars , pch=16, las=1,xlab="logMeanExpression",ylab="variance",
      cex=ifelse( rownames(trx) %in% top_variable , .2,.1),
      col= ifelse( rownames(trx) %in% top_variable , "red","black"))

# Identify common samples between the transcriptome and the mirobiome dataset
common_samples <- colnames(microbiome)[colnames(microbiome) %in% colnames(trx)]
trx <- trx[,as.character(common_samples) ]
microbiome2 <- microbiome[,as.character(common_samples)]

# Combine datased and set all to same scale using Z-score transformation
mypar(1,3)
merged <- rbind(trx[top_variable,],microbiome2)
zscore_merged <- t(apply(merged,1,function(i){scale(i,T,T)}))
zscore_merged[ zscore_merged > 5 ] <- 5
zscore_merged[ zscore_merged < -5 ] <- -5
rownames(zscore_merged) <- rownames(zscore_merged)
colnames(zscore_merged) <- colnames(zscore_merged)
hist(zscore_merged,breaks = 100,main="all")
hist(zscore_merged[top_variable,],breaks = 100,main="trx")
hist(zscore_merged[rownames(microbiome2),],breaks = 100,main="microbiome")

PC <- prcomp(zscore_merged,center = F,scale. = F)
plot(PC$sdev,pch=21,bg="grey70")
abline(v=50.5,col="red",xpd=F,lty=2)

knn <- RANN::nn2( query = PC$x[rownames(microbiome2),1:50] , 
                data = PC$x[top_variable,1:50] ,k = 50)
# knn <- RANN::nn2(t(zscore),k = k)
nn <- matrix(0, length(rownames(microbiome2)), length(top_variable))
for(idx in 1:nrow(knn$nn.idx) ){  nn[idx,knn$nn.idx[idx,]] <- 1 }
nn <- nn[ rowSums(nn)>0 , colSums(nn)>0 ]

mypar(1,2,mar=c(0,0,2,10))

g <- graph_from_adjacency_matrix(nn, mode = "undirected",diag = F,weighted = T)
set.seed(1)
# l <- layout_with_graphopt(g,niter = 1000)
l <- layout_with_fr(g,niter=500,start.temp=100)

```{r, eval=F,echo = FALSE, fig.width  = 10, fig.height=11}

>>>>

---
title: "`r params$project_title`"
editor_options:
  chunk_output_type: console
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
  html_document:
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: no
params:
  dataset: 1
---

@@

#Graph construction  
fct <- .25 # FC threshold for differential expression
pvt <- 0.05 # Pvalue threshold for differential expression
min_pct <- .1 # minimun level of detected bacteria in each sample group

#dataset use
dataset_use <- 3
```

@@

@@

#import datasets and return matrix with taxa-level resolution
datasets <- lapply(dataset_names, function(x){
  x <- read.csv( paste0(PATH,"/results/",x),row.names = 1)
  return(x)
})
names(datasets) <- dataset_names
lapply(datasets,dim)
```

@@

# Create empty matrix for the full dataset
microbiome <- as.matrix(datasets[[ params$dataset ]])
microbiome <- microbiome[ rowSums(microbiome>0)>=2 , colSums(microbiome>0)>=2 ]

@@

# Saving clusters and differentially expressed bacteria

```{r, eval=T,echo = FALSE}
write.csv2( data.frame(cl$membership,row.names = colnames(microbiome)) ,row.names = T,
            paste0(PATH,"/results/Sample_Louvain_clusters_",params$dataset) )

write.csv2( data.frame(l,row.names = colnames(microbiome)) ,row.names = T,
            paste0(PATH,"/results/Sample_graph_layout_",params$dataset) )

write.csv2( res ,
            paste0(PATH,"/results/DGE_",params$dataset),row.names = T )
```

### 2020-10-28 reports/rmarkdown/02b_evaluating_batch_correction.Rmd
adds the report for comparing microbiome datasets

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(fig.width  = 10,
                      results    = "hold",
                      message    = FALSE,
                      warning    = FALSE)
```

@@

```{r, echo = FALSE}
library(igraph)
library(rafalib)
library(sva)
library(batchelor)
source("~/repos/niceRplots/R/plotting_functions.R")
source("~/repos/niceRplots/R/add_fig_label.R")
source("~/repos/niceRplots/R/helper_functions.R")

@@

```{r, echo = FALSE}
PATH <- "~/Desktop/NBIS/SMS_Projects/broliden_5325/" #Path to the data

@@

```{r, echo = FALSE}
# Load metadata
metadata <- read.csv(paste0(PATH,"/data/Clinical_visit_2_3_updatesept28.csv"))
rownames(metadata) <- as.character(metadata$ID)

@@

# Calculate QC metrics

```{r, echo = FALSE}
counts <- t(as.data.frame( lapply(datasets_all_samples,colSums) ))
counts[is.na(counts)] <- 0
detected_counts <- t(as.data.frame( lapply(datasets_all_samples,function(x) colSums( x > 0 )) ))
detected_counts[is.na(detected_counts)] <- 0
rownames(detected_counts) <- paste0(rownames(detected_counts),">0")

mypar(mar=c(2,20,2,1))
barlist( data = rbind( counts , detected_counts),
         genes = c(rownames(counts),rownames(detected_counts)),
         clustering = rep(1,nrow(metadata)),
         col=ifelse( (colSums(counts == 0)>=3)*1,"red","grey" ),draw_mean_lines=F)

sum((colSums(counts == 0)>=3)*1)
colnames(counts)[(colSums(counts == 0)>=3)]

metadata[colnames(counts)[(colSums(counts == 0)>=3)],]
```

@@

```{r, echo = FALSE}
mypar(1,2,mar=c(2,15,1,1))

@@

```{r, echo = FALSE}
# Computing differential expression across datasets

all_microbiome <- cbind(datasets_all_samples[["ASV_tissue_V3_normalized_batch_corrected"]],
                        datasets_all_samples[["ASV_CVL_V3_normalized_batch_corrected"]],
                        datasets_all_samples[["ASV_CVL_V2_normalized_batch_corrected"]])
datasets <- factor(c( rep("tissue_v3",121) , rep("CVL_v3",121) , rep("CVL_v2",121) ))

    temp <- wilcox.test(x=a, y=b)

    res <- rbind(res, setNames(c(j,i,fc,perc1,perc2,unlist(temp)[2] ),
                          c("bacteria","cluster","fc","perc.1","perc.2","pvalue")) )
    colnames(res) <- c("bacteria","cluster","fc","perc.1","perc.2","pvalue")
  }
}
res <- res[-1,]
res$pvalue <- as.numeric(res$pvalue)
res$perc.1 <- as.numeric(res$perc.1)
res$perc.2 <- as.numeric(res$perc.2)
res$perc.diff <- res$perc.1 - res$perc.2

res$fc <- as.numeric(res$fc)
res$FDR <- p.adjust(res$pvalue)
res <- res[order(res$pvalue),]
res <- res[abs(res$fc) > fct,]
res <- res[abs(res$pvalue) < pvt,]
res <- res[ (res$perc.1 > min_pct) | (res$perc.2 > min_pct) ,]
dim(res)

mypar(mar=c(2,15,1,1))
plot_dots( all_microbiome , names(sort(ord)) , clustering = datasets,show_grid = T,cex.main=1,font.main=1,
          cex.row = .8,cex.col = .8,srt = 0)

# Correlation CVL3 and RNAseq

```{r, echo = FALSE}

TRX <- datasets_all_samples[["Tissue_RNAseq_V3_normalized"]]
CVL3 <- datasets_all_samples[["ASV_CVL_V3_normalized_batch_corrected"]]
samples_TRX_and_CVL3 <- colnames(counts) [ colSums(counts[c("Tissue_RNAseq_V3_normalized","ASV_CVL_V3_normalized_batch_corrected"),] > 0)==2 ]

#filter TRX dataset
TRX <- TRX [ , samples_TRX_and_CVL3 ]
TRX <- TRX [ rowSums(TRX>0)>= 2 , ]
top_vars_TRX <- names(sort(apply(TRX,1,var),decreasing = T)[1:3000] )
plot(apply(TRX,1,mean),apply(TRX,1,var),col=ifelse(rownames(TRX) %in% top_vars_TRX,"red","grey20"),cex=.2)
TRX <- TRX[ top_vars_TRX , ]
dim(TRX)

#filter CVL3 dataset
CVL3 <- CVL3 [ , samples_TRX_and_CVL3 ]
CVL3 <- CVL3 [ rowSums(CVL3>0)>= 2 , ]
top_vars_CVL3 <- names(sort(apply(CVL3,1,var),decreasing = T)[1:100] )
dim(CVL3)

mypar()
cors <- cor( t(rbind(TRX) ) ,  t(rbind(CVL3) ) )
hist(cors,xlim=c(-1,1),breaks = 100)
dim(cors)

library(fgsea)
bacteria_use <- colnames(cors)
gmt <- gmtPathways("../../supplementary_files/c2.cp.kegg.v6.2.symbols.gmt.txt")
enrichments <- lapply(bacteria_use,function(x){
  res <- fgsea(pathways = gmt,stats = sort(cors[,x],decreasing = T),nper=1000)
  return(res)
})
names(enrichments) <- bacteria_use

pvalues <- lapply(enrichments,function(x) setNames(x$pval,x$pathway) )
pvalues <- as.data.frame(pvalues)
pvalues <- -log10( pvalues )
top_pathways <- names(sort(apply(pvalues,1,median),T))[1:50]
mypar(mar=c(4,15,2,2))
boxplot( t(pvalues[rev(top_pathways),]),cex=.1,horizontal=T,las=1 )

pvalues[ pvalues < -log10(0.05)] <- 0
sort(rowSums(pvalues>0),T)
sort(colSums(pvalues>0),T)

NESes <- lapply(enrichments,function(x) setNames(x$NES,x$pathway) )
NESes <- t(as.data.frame(NESes))
# 
# o_kegg <- hclust( as.dist( (1-cor(NESes))/2 ),"ward.D2")$order
# o_bacs <- hclust( as.dist( (1-cor(t(NESes)))/2 ),"ward.D2")$order
o_kegg <- hclust( dist( t(NESes) ),"ward.D2")$order
o_bacs <- hclust( dist( NESes ),"ward.D2")$order
NESes <- NESes[o_bacs,o_kegg]

mypar(mar=c(15,2,2,15))
image( t(NESes[nrow(NESes):1,]),col=colorRampPalette(c("navy","grey95","firebrick"))(91),breaks=seq(-4,4,length.out = 92),
       axes=F,border=NA,main="KEGG pathways",xlab="",ylab="bacterias",line=.4,cex.main=1,font.main=1)
text(  par("usr")[c(4)]+.02 , seq(1,0,length.out = nrow(NESes)),
      labels = rownames(NESes), srt = 0, adj = c(0,.5), xpd = TRUE, cex=.3)
text( seq(0,1,length.out = nrow(NESes)) , par("usr")[c(1)]-.02,
      labels = colnames(NESes), srt = 90, adj = c(1,.5), xpd = TRUE, cex=.3)

eee <- data.frame(x=c(sapply(colnames(NESes),function(x)rep(x,nrow(NESes)))),
                  y=rep(rownames(NESes),ncol(NESes)),
                  w=c(NESes),
                  p=c(t(pvalues)),
                  x1=0,
                  y1=c(sapply(ncol(NESes):1,function(x)rep(x,nrow(NESes))))/ncol(NESes) ,
                  x2=1,
                  y2=rep(nrow(NESes):1,ncol(NESes))/nrow(NESes) )

eee$w2 <- eee$w / max(eee$w)
eee <- eee[eee$w > 1.5,]
eee <- eee[eee$p > 2,]

eee$y1 <- rank(eee$y1) / length(eee$y1)
eee$y2 <- rank(eee$y2) / length(eee$y2)

plot(c(0,1),c(0,1),type="n")
for( i in 1:nrow(eee)){
  lines(c(eee$x1[i],eee$x2[i]),c(eee$y1[i],eee$y2[i]),
        col=colorRampPalette(c("grey95","grey","firebrick"))(99)[ eee$w2 * 98 + 1 ] )
}

```{r, echo = FALSE}

g <- graph_from_incidence_matrix( 1-cors, directed = F,weighted = T  )

l <- layout_with_fr(g)
plot( g , vertex.label.cex=0.000001 , layout=l,
      edge.width=  ( E(g)$weight / max(E(g)$weight)) ,
      vertex.size=1,vertex.color=ifelse(V(g)$name %in% colnames(CVL3),"red","blue") )

>>>>

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(fig.width  = 10,
                      results    = "hide",
                      message    = FALSE,
                      warning    = FALSE)
```

@@

```{r}
library(igraph)
library(rafalib)
library(sva)
library(batchelor)
source("~/repos/niceRplots/R/plotting_functions.R")
source("~/repos/niceRplots/R/add_fig_label.R")
source("~/repos/niceRplots/R/helper_functions.R")

@@

```{r}
PATH <- "~/Desktop/NBIS/SMS_Projects/broliden_5325/" #Path to the data

@@

```{r}
# Load metadata
metadata <- read.csv(paste0(PATH,"/data/Clinical_visit_2_3_updatesept28.csv"))
rownames(metadata) <- as.character(metadata$ID)

@@

@@

```{r}
mypar(1,2,mar=c(2,15,1,1))

@@
